---
title: "Analyzing S&P 500 Drawdowns"
author: "Arkadi Avanesyan"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Introduction

Understanding historical drawdowns is essential for building crisis-proof portfolios.
In this tutorial, we'll fetch S&P 500 data and identify every major market crash since 1990
using the `PerformanceAnalytics` package.

## 1. Install required packages

```{r install, eval=FALSE}
install.packages(c("quantmod", "PerformanceAnalytics"))
```

## 2. Fetch S&P 500 data since 1990

```{r fetch-data}
library(quantmod)
library(PerformanceAnalytics)

# Fetch S&P 500 data from 1990
getSymbols("^GSPC", src = "yahoo", from = "1990-01-01")

# Calculate daily returns from adjusted close
sp500_returns <- dailyReturn(Ad(GSPC))
colnames(sp500_returns) <- "SP500"

# Quick summary
head(sp500_returns)
tail(sp500_returns)
```

## 3. Visualize the drawdown history

```{r drawdown-chart, fig.width=10, fig.height=6}
# Plot drawdown chart
chart.Drawdown(sp500_returns,
               main = "S&P 500 Drawdowns Since 1990",
               colorset = "darkred")
```

This shows every period where the market was below its previous peak.
The deeper the red, the worse the drawdown.

## 4. Extract the worst drawdowns

```{r worst-drawdowns}
# Get table of worst drawdowns
drawdowns <- table.Drawdowns(sp500_returns, top = 10)
print(drawdowns)
```

This returns:
- **From**: Start date of the drawdown
- **Trough**: Date of the bottom
- **To**: Recovery date
- **Depth**: Maximum loss
- **Length**: Days to reach bottom
- **Recovery**: Total days underwater

## 5. Calculate key drawdown statistics

```{r statistics}
# Maximum drawdown
cat("Maximum Drawdown:", maxDrawdown(sp500_returns), "\n")

# Average drawdown
cat("Average Drawdown:", AverageDrawdown(sp500_returns), "\n")

# Calmar ratio (return / max drawdown)
cat("Calmar Ratio:", CalmarRatio(sp500_returns), "\n")

# Pain index (average depth * duration)
cat("Pain Index:", PainIndex(sp500_returns), "\n")
```

## 6. Identify crisis periods

```{r major-crashes}
# Get all drawdowns deeper than 20%
all_drawdowns <- table.Drawdowns(sp500_returns, top = 20)
major_crashes <- all_drawdowns[all_drawdowns$Depth < -0.20, ]
print(major_crashes)
```

### Major crashes you'll typically see:
- **2008 Financial Crisis**: ~55% drawdown
- **2020 COVID Crash**: ~34% drawdown
- **2000-2002 Dot-com Bust**: ~49% drawdown
- **2022 Bear Market**: ~25% drawdown

## Key Insight

The S&P 500 has experienced drawdowns exceeding 20% roughly once every 5-7 years.
A crisis-proof portfolio should be designed to either survive these events or benefit from them.
This is the foundation of antifragile portfolio construction.

---

**Learn more**: Join my free workshop on building crisis-proof portfolios at
[aavanesy.github.io/workshops](https://aavanesy.github.io/workshops)
